Vegetation regrowth/encroachment (gjengroing)

### 2023-08-24 DRAFT_functional_plant_indicators_natopen.Rmd
starting the natopen markdown

# Functional plant indicators, semi-natural ecosystems (Light, Moisture, pH, Nitrogen, Phosphorus, Grazing_mowing, Soil disturbance) {#Functional-plant-indicators-seminat}

@@

May 2023

@@

```{r, echo=FALSE}
Ecosystem <- "semi-natural"
Egenskap  <- c("Primærproduksjon","Abiotiske forhold")
ECT       <- "Functional state characteristic"
Contact   <- "Joachim Töpper"
```

@@

## Introduction
Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003). The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf. Töpper et al. 2018). For instance, using an indicator for moisture one would find a functional signature with higher moisture values for plant communities in mires compared to e.g. grasslands or forests. Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition. Here, we combine functional plant indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al. 2019) for semi-natural ecosystems. We calculate the functional signature of plant communities in the monitored sites with respect to light, moisture, pH, nitrogen, phosphorus, grazing_mowing, and soil disturbance. These functional signatures are then compared to reference distributions of functional signature, separately for each wetland ecosystem type, calculated from 'generalized species lists' underlying the Norwegian categorization system for eco-diversity (Halvorsen et al. 2020). These plant functional condition indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al. 2021, Töpper & Jakobsson 2021). Note that deviations from the reference may occur in both directions, e.g. the moisture signature from the testing data may be higher or lower than in the reference. Deviations in these two directions indicate very different environmental phenomena and thus have to be treated separately. Therefore, we develop two condition indicators for each functional plant indicator, a lower one and an upper one (see further down for more details).

## About the underlying data
In the 'functional plant indicator' project for semi-natural ecosystems, we use four sets of data for building indicators for ecological condition:

- as test data we use plant community data from (1) the ANO monitoring scheme (cf. Tingstad et al. 2019, contains semi-natural grasslands(T32), semi-natural wet meadows (V10), heathlands (T34), and we also decided to include near-semi-natural meadows (T41) ) and (2) from the ASO monitoring scheme (cf. Bär et al. 2021, targets semi-natural grasslands (T32) only)
- as reference data we use (3) generalized species lists from the Norwegian categorization system for eco-diversity (NiN) (cf. Halvorsen et al. 2020)
- (4) Swedish plant indicator data for light, moisture, pH, and nitrogen from Tyler et al. (2021)

@@

(2) ASO monitoring data
ASO stands for 'arealrepresentativ overvåking av semi-naturlig eng', i.e. 'area representative monitoring of semi-natural meadows'. In ASO ca. 20 areas are chosen annually based on a probabilistic distribution model of semi-natural meadows in Norway (see Johansen et al. 2017 for details). Each ASO area (område) spans a 10 x 10 km grid cell, in which a maximum of up to 10 sites (flate) of 500 x 500 m are chosen. In each site, all semi-natural meadows (eng) with an area of at least 250 sqm are delineated. In up to three meadows per site the plant species composition is sampled using a transect method. For the vegetation transect analyses, both cover and frequency of each vascular plant species along the transect is recorded. For every meadow, the species data collection is accompanied by an assessment of the ecosystem, including ecosystem type and some additional variables related to ecosystem-specific drivers of state.
In the analysis in this document, we use the ASO data from 2022 only as the data from 2021 and the ASO pilot in 2019 are not in sufficiently good shape. In the future, all ASO data will hopefully be available in Miljødirektoratets kartkatalog.

@@

## References
Bär, A., Albertsen, E., Bele, B., Daugstad, K., Grenne, S.N., Jakobsson, S., Solbu, E.B., Thorvaldsen, P., Vesterbukt, P., Wehn, S. & Johansen, L. 2021. Utvikling av nasjonal arealrepresentativ overvåking av semi-naturlig eng (ASO). Uttesting, ferdigstilling og utvalg av områder. NIBIO Rapport 7 (7). 84 s.

Diekmann, M. 2003. Species indicator values as an important tool in applied plant ecology - a review. Basic and Applied Ecology 4: 493-506, doi:10.1078/1439-1791-00185

@@

Johansen, L., S. Wehn, R. Halvorsen, og K. Hovstad. 2017. Metode for overvåking av semi-naturlig eng i Norge. NIBIO Rapport 3 (25). 35 s.

>>>>

# Functional plant indicators, naturally open ecosystems below the tree line (Grime's CSR values, Light, Nitrogen, Soil disturbance) {#Functional-plant-indicators-seminat}

@@

August 2023

@@

```{r, echo=FALSE}
Ecosystem <- "naturally open below tree line"
Egenskap  <- c("Primærproduksjon","Abiotiske forhold")
ECT       <- "Functional state characteristic"
Contact   <- "Joachim Töpper"
```

@@

## Introduction
Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003). The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf. Töpper et al. 2018). For instance, using an indicator for moisture one would find a functional signature with higher moisture values for plant communities in mires compared to e.g. grasslands or forests. Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition. Here, we combine functional plant indicator data with field sampled plant community data from the Norwegian nature monitoring programs ANO (Tingstad et al. 2019) and GRUK (Evju et al. 2020) for naturally open ecosystems below tree line (abbreviated as 'nat-open' henceforth). We calculate the functional signature of plant communities in the monitored sites with respect to Grime's CSR values, light, nitrogen, and soil disturbance. These functional signatures are then compared to reference distributions of functional signature, separately for each nat-open ecosystem type, calculated from 'generalized species lists' developed for ecosystem types in the Norwegian categorization system for eco-diversity (Halvorsen et al. 2020). These plant functional condition indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al. 2021, Töpper & Jakobsson 2021). Note that deviations from the reference may occur in both directions, e.g. the nitrogen signature from the testing data may be higher or lower than in the reference. Deviations in these two directions indicate very different environmental phenomena and thus have to be treated separately. Therefore, we develop two condition indicators for each functional plant indicator, a lower one and an upper one (see further down for more details).

## About the underlying data
In the 'functional plant indicator' project for nat-open ecosystems, we use four sets of data for building indicators for ecological condition:

- as test data we use plant community data (1) from the ANO monitoring scheme (cf. Tingstad et al. 2019 for the years 2019-2021, which among many other ecosystem types contains the nat-open types "Åpen grunnlendt mark" (T2), coastal meadows (T12), "meadows and heath in rock slides"Rasmarkhei og -eng" (T16), and "we also decided to include near-semi-natural meadows"Åpen flomfastmark" (T18) ) and (2) from the GRUK monitoring scheme (cf. Evju et al. 2020), which focuses on limestone-rich "Åpen grunnlendt mark" (T2) around the Oslofjord.
- as reference data we use (3) generalized species lists developed by Evju et al. 2023 for nat-open ecosystem types.
- (4) Swedish plant indicator data for light, moisture, pH, and nitrogen from Tyler et al. (2021)

@@

#### continue here ####

(2) GRUK monitoring data
ASO stands for 'arealrepresentativ overvåking av semi-naturlig eng', i.e. 'area representative monitoring of semi-natural meadows'. In ASO ca. 20 areas are chosen annually based on a probabilistic distribution model of semi-natural meadows in Norway (see Johansen et al. 2017 for details). Each ASO area (område) spans a 10 x 10 km grid cell, in which a maximum of up to 10 sites (flate) of 500 x 500 m are chosen. In each site, all semi-natural meadows (eng) with an area of at least 250 sqm are delineated. In up to three meadows per site the plant species composition is sampled using a transect method. For the vegetation transect analyses, both cover and frequency of each vascular plant species along the transect is recorded. For every meadow, the species data collection is accompanied by an assessment of the ecosystem, including ecosystem type and some additional variables related to ecosystem-specific drivers of state.
In the analysis in this document, we use the ASO data from 2022 only as the data from 2021 and the ASO pilot in 2019 are not in sufficiently good shape. In the future, all ASO data will hopefully be available in Miljødirektoratets kartkatalog.

@@

## References
Diekmann, M. 2003. Species indicator values as an important tool in applied plant ecology - a review. Basic and Applied Ecology 4: 493-506, doi:10.1078/1439-1791-00185

Evju, M., Stabbetorp, O.E., Olsen, S.L., Bratli, H., Often, A. & Bakkestuen, V. 2020. Dry calcareous grasslands in the Oslofjord region. A test of monitoring protocols and results for 2020. NINA Report 1910. Norwegian Institute for Nature Research.

Evju, M., Stabbetorp, O.E., Olsen, S.L., Bratli, ... . 2023. Generalized species lists for naturally open ecosystem types below the tree line in Norway. in prep.

@@
